Bridging Gaps: Investigating COVID-19's Influence on Health Disparities in Connecticut

### 2023-12-16 manuscript/manuscript.tex
replaced by year graph with new format

Income is a fundamental
social need that significantly influences
one's health. The level of income
directly affects an individual's access
to essential resources such as nutritious
food, healthcare, and secure housing.
People with lower incomes often face
barriers to healthcare services and
struggle to afford a balanced diet,
leading to higher risks of chronic
illnesses. Income disparities contribute
to inequalities in health outcomes, as
those with limited financial resources
may experience higher stress levels and
reduced access to preventive care. 
Data on income and mortality from the National Longitudinal Mortality Study reported by Backlund, Sorlie, and Johnson (1999) (4)
reveals that a continued drop in mortality results as income increases with a more significant drop towards the lower portion of the distribution(4).
Addressing income inequality is crucial
for promoting overall well-being and
reducing health disparities in communities.

Education is a key social determinant
of health, influencing various aspects
of an individual's life. Higher levels
of education are associated with better
health outcomes, as education enhances
one's ability to make informed decisions
about lifestyle choices, healthcare
utilization, and disease prevention. In CT, it is revealed there has been a spike in chronic absenteeism during the onset of the pandemic.
''Specifically, between the 2018-19 and 2021-22 school years, the share of students chronically absent grew by 13.5 percentage points.''(1)
There is also evidence of decreased Connecticut Performance Index (CPI) below the state target levels from 2019 to 2022 in mathematics, science,
and language arts.(1)
Limited educational opportunities can
result in lower health literacy, reducing
an individual's capacity to navigate
complex healthcare systems and understand
preventive measures. Educational
disparities contribute to a cycle of
poor health outcomes, as individuals
with lower levels of education may face
challenges in securing stable employment
and accessing adequate healthcare
resources. COVID-19 has only exaggerated
the pre-existing opportunity gap between urban schools and suburban schools in providing equal educational
opportunities despite both being in Connecticut.(2)
From (5), data reveals that  eath rates are lower for those with more education for all-cause mortality. 
The gap between high school graduates and those with some post-secondary education are larger compared to the latter group consisting
of a mixture of secondary education including college and advanced degrees.

Access to sufficient
food is a critical social need that
directly impacts health. Food insecurity,
often linked to low income, can lead to
malnutrition, obesity, and various health
issues. Lack of access to fresh and
healthy foods contributes to the prevalence
of chronic conditions such as diabetes
and cardiovascular diseases. Current
issues related to food include disparities
in food availability, affordability, and
nutritional quality, disproportionately
affecting marginalized communities.
Addressing these issues involves promoting
food equity, supporting local agriculture,
and implementing policies that ensure
everyone has access to nutritious food,
ultimately improving overall public health.

Housing plays a vital role in determining
health outcomes. Inadequate or unstable
housing conditions can lead to physical
and mental health issues. Homelessness
and substandard living conditions expose
individuals to environmental hazards,
increase stress levels, and contribute
to the spread of infectious diseases.
The current housing crisis in many regions
exacerbates these issues, with rising
costs and limited affordable housing
options. Addressing housing as a social
need involves implementing policies to
ensure affordable housing, preventing
homelessness, and improving living
conditions to create a foundation for
better health outcomes.

Access to healthcare through insurance
is a critical social need that significantly
impacts health outcomes. Lack of health
insurance can lead to delayed or forgone
medical care, resulting in the progression
of illnesses and poorer health.
Disparities in insurance coverage contribute
to inequalities in healthcare access and
health outcomes. Current issues include
the high costs of healthcare and the
existence of uninsured or underinsured
populations. Addressing these challenges
involves expanding access to affordable
health insurance, implementing healthcare
reforms, and promoting policies that
ensure comprehensive coverage for all
individuals.

Rehospitalization refers to the phenomenon
where individuals experience repeated
hospital admissions, often due to
complications or inadequate follow-up care.
High rates of rehospitalization indicate
gaps in healthcare delivery and management,
impacting both individual well-being and
healthcare system efficiency. This social
need is indicative of issues such as
insufficient post-discharge support,
medication management, and coordination
of care. Repeated hospitalizations not
only strain healthcare resources but also
have significant implications for the
individuals involved, leading to increased
morbidity, reduced quality of life, and
higher healthcare costs. Addressing
rehospitalization requires improvements
in transitional care, enhanced coordination
among healthcare providers, and better
support systems for patients, especially
those with chronic conditions, to ensure
a smoother recovery and reduce the burden
on healthcare systems. Recognizing and
addressing these factors can contribute
to overall health improvement, prevent
unnecessary healthcare utilization, and
enhance the effectiveness of healthcare
services.

@@

In this study, descriptive statistics is utilized to outline the total population, racial composition, 
and education levels across the 8 counties in Connecticut. ANOVA analyses were 
conducted to assess the significance of the difference of the variables of median income, health insurance, food access, housing, and rehospitalization rates between counties and across the four years. Additional 
Tukey's HSD tests were conducted to determine the specific counties and years that have had significant differences within 
each of the variables for each county.

@@

  \begin{figure}[tbp]
    \label{fig:Social Needs by YearGraph}
      \centering
      \includegraphics[width=\textwidth]{By Year.pdf}
      \caption{Social Needs Multiple Comparisons by County.}
    \end{figure}

>>>>

The level of income
directly affects an individual's access
to essential resources such as nutritious
food, healthcare, and secure housing.
Those with lower incomes often face
barriers to healthcare services and
struggle to afford a balanced diet,
leading to higher risks of chronic
illnesses. Income disparities contribute
to inequalities in health outcomes, as
those with limited financial resources
may experience higher stress levels and
reduced access to preventive care. 
Data on income and mortality from the National Longitudinal Mortality Study reported by Backlund, Sorlie, and Johnson (1999) (4)
reveals that a continued drop in mortality results as income increases with a more significant drop towards the lower portion of the distribution(4).
Addressing income inequality is crucial
for promoting overall well-being and
reducing health disparities in communities.

Higher levels
of education are associated with better
health outcomes, as education enhances
one's ability to make informed decisions
about lifestyle choices, healthcare
utilization, and disease prevention. In CT, it is revealed there has been a spike in chronic absenteeism during the onset of the pandemic.
''Specifically, between the 2018-19 and 2021-22 school years, the share of students chronically absent grew by 13.5 percentage points.''(1)
There is also evidence of decreased Connecticut Performance Index (CPI) below the state target levels from 2019 to 2022 in mathematics, science,
and language arts.(1)
Limited educational opportunities can
result in lower health literacy, reducing
an individual's capacity to navigate
complex healthcare systems and understand
preventive measures. Educational
disparities contribute to a cycle of
poor health outcomes, as individuals
with lower levels of education may face
challenges in securing stable employment
and accessing adequate healthcare
resources. COVID-19 has only exaggerated
the pre-existing opportunity gap between urban schools and suburban schools in providing equal educational
opportunities despite both being in Connecticut.(2)
From (5), data reveals that  eath rates are lower for those with more education for all-cause mortality. 
The gap between high school graduates and those with some post-secondary education are larger compared to the latter group consisting
of a mixture of secondary education including college and advanced degrees.

Food insecurity,
often linked to low income, can lead to
malnutrition, obesity, and various health
issues. Lack of access to fresh and
healthy foods contributes to the prevalence
of chronic conditions such as diabetes
and cardiovascular diseases. Current
issues related to food include disparities
in food availability, affordability, and
nutritional quality, disproportionately
affecting marginalized communities.
In 2009, a nationwide analysis conducted by the USDA underscored the significance of both income inequality and racial segregation(6).
The analysis indicated that in low-income areas, corner stores and similar establishments often provide 
a surplus of processed foods with limited access to fresh items. However, healthier food options are 
typically more prevalent in predominantly white communities compared to non-white communities.
Addressing these issues involves promoting
food equity, supporting local agriculture,
and implementing policies that ensure
everyone has access to nutritious food,
ultimately improving overall public health.

Housing plays a vital role in determining
health outcomes. Inadequate or unstable
housing conditions can lead to physical
and mental health issues. Homelessness
and substandard living conditions expose
individuals to environmental hazards,
increase stress levels, and contribute
to the spread of infectious diseases.
Recent literature(8) has assessed the relationship between outcomes of health 
and housing factors of physical health; mental health; chronic and infectious disease; maternal,
 infant, reproductive, and sexual health; and injury. Studies have revealed that low income 
 populations and those with high proportions of racial and ethnic minority groups face greater exposures to environmental hazards(8)
Four pillars of housing has been used as a framework for assessing health-related aspects of 
housing(7): 1) conditions (quality of housing); 2) cost (affordability); 3) consistency (residential stability); 
and 4) context (neighborhood opportunities).
The current housing crisis in many regions
exacerbates these issues, with rising
costs and limited affordable housing
options. Addressing housing as a social
need involves implementing policies to
ensure affordable housing, preventing
homelessness, and improving living
conditions to create a foundation for
better health outcomes.

Access to healthcare through insurance
is a critical social need that significantly
impacts health outcomes. Lack of health
insurance can lead to delayed or forgone
medical care, resulting in the progression
of illnesses and poorer health. 
Large disparities in health insurance coverage related to race and ethnicity remain a persisting issue of the US health care system
further exacerbating health inequalities(9).
Disparities in insurance coverage contribute
to inequalities in healthcare access and
health outcomes. Current issues include
the high costs of healthcare and the
existence of uninsured or underinsured
populations. Addressing these challenges
involves expanding access to affordable
health insurance, implementing healthcare
reforms, and promoting policies that
ensure comprehensive coverage for all
individuals.

Rehospitalization refers to the phenomenon
where individuals experience repeated
hospital admissions, often due to
complications or inadequate follow-up care.
High rates of rehospitalization indicate
gaps in healthcare delivery and management,
impacting both individual well-being and
healthcare system efficiency(9).

In 2011, around 3.3 million adults experienced hospital readmissions, 
incurring associated costs amounting to approximately \$41.3 billion(10).

This social
need is indicative of issues such as
insufficient post-discharge support,
medication management, and coordination
of care. Repeated hospitalizations not
only strain healthcare resources but also
have significant implications for the
individuals involved, leading to increased
morbidity, reduced quality of life, and
higher healthcare costs(9). Addressing
rehospitalization requires improvements
in transitional care, enhanced coordination
among healthcare providers, and better
support systems for patients. Recognizing and
addressing these factors can contribute
to overall health improvement, prevent
unnecessary healthcare utilization, and
enhance the effectiveness of healthcare
services.

@@

In this study, descriptive statistics is utilized to outline the total population, racial composition, 
and education levels across the 8 counties in Connecticut. ANOVA analyses were 
conducted to assess the significance of the difference of the variables of median income, health insurance, food access, housing, 
and rehospitalization rates between counties and across the four years using SAS v.9.4. Additional 
Tukey's HSD tests were conducted to determine the specific counties and years that have had significant differences within 
each of the variables for each county using SAS v.9.4 and SPSS v.29.

@@

  \begin{figure}[tbp]
    \label{fig:Social Needs by YearGraph}
      \centering
      \includegraphics[width=\textwidth]{Social Need by Year.pdf}
      \caption{Social Needs Multiple Comparisons by County.}
    \end{figure}
